\chapter{June 26, 2024}

\section{Functions in Multiple Variables}

On top of vector functions and standard single-input, single-ouput functions, there are functions that can take multiple inputs (and produce multiple outputs, but that's not needed right now). These are called multivariable functions. We start by looking at the domain of such functions.

\begin{example}
    Find the domain of
    \[f(x, y) = \frac{\sqrt{x^{2} + y^{2}}}{x - y}.\]

    \begin{soln}
        Now, finding domains in multiple variables often isn`t much more difficult than finding domains in one variable. Functions are still defined according to the same mathematical laws. As such, the denominator cannot equal to zero. So, the domain is
        \[D_{f} = \{(x, y) \in \RR^{2} \mid x \neq y\}.\]
    \end{soln}
\end{example}

Just as limits are defined with single-input functions, they are also useful for multi-input functions. However, they require a bit more than the standard limits we learned in earlier classes.

\begin{definition}
    Let $f : \RR^{2} \to \RR$ be a function of two variables. We say
    \[\lim_{(x, y) \to (x_{0}, y_{0})} f(x, y) = L\]
    if and only if for any $\epsilon > 0$ there exists $\delta > 0$ such that
    \[\abs{f(x, y) - L} < \epsilon\]
    whenever
    \[0 < (x - x_{0})^{2} + (y - y_{0})^{2} < \delta.\]
\end{definition}

Behind the complicated notation, the definition is quite similar to the definition of a limit in $\RR^{2}$. The main difference is instead of the function approaching a value from the negative and positive direction, it must approach the value from a disk of points around the desired limit point.

The addition of extra variables also makes evaluating these limits a bit different. Albeit, there is still a well-defined and straightforward way to do so in many cases and many techniques from a single variable carry over. The below example explores a technique that arises only with a second variable.

\begin{example}
    Find
    \[\lim_{(x, y) \to (0, 0)} \frac{\sin\left(x^{2} + y^{2}\right)}{x^{2} + y^{2}}.\]

    \begin{soln}
        This solution involves the technique of converting frames to polar coordinates that is neither needed nor useful for single variable limits. By the definition of the transition we set
        \[x^{2} + y^{2} = r^{2}\]
        and subsitute to yield
        \[\lim_{r \to 0} \frac{\sin r^{2}}{r^{2}}.\]
        Now, by l`Hopital's rule our limit simplifies to
        \[\lim_{r \to 0} \frac{2r \cos r^{2}}{2r} = 1.\]
    \end{soln}
\end{example}

\begin{remark}
    In the above limit we make use of l'Hopital's rule, which has no simple direct application to multiple variables. However, in cases like the one above when the limit can be reduced to a single variable (either through polar coordinates, searching along a path, or any other way), l'Hopital's rule is both applicable and incredibly useful.
\end{remark}

A substantial way to prove that a limit does not exist comes in the form of searching for limits along a path.

\begin{example}
    Find
    \[\lim_{(x, y) \to (0, 0)} \frac{y^{2}}{x^{2} + y^{2}}.\]

    \begin{soln}
        After toying around with the limit for a while, you may find it difficult to approach a solution. So, let's evaluate the limit along a path. First, consider a modification of the function, which we denote $f(x, y)$, to $f(0, y)$. Along this path of $x = 0$, we see that
        \[f(0, y) \to 1 \; \text{as} \; (x, y) \to (0, 0)\]
        via a simplification of the fraction. However, in the case where we analyze $f(x, 0)$, we find that
        \[f(x, 0) \to 0 \; \text{as} \; (x, y) \to (0, 0).\]
        So, our previously stated definition of a limit fails and we conclude that the limit does not exist.
    \end{soln}
\end{example}

\begin{remark}
    The technique of path following is a nice heuristic for proving a limit does not exist. However, it cannot be used to prove the existance of a limit. Just because you check some paths and they happen to approach the same limit $L$ at a point $(x_{0}, y_{0})$, does not mean that other paths and directions approach the same point. The existance of a limit cannot be verified until it is proved in its entirety.
\end{remark}